---
title: "Oak Data Exploration"
format: html
editor: visual
---

## Oak Exploratory data analysis and cleaning

```{r}
#install.packages("tidyverse")
#install.packages("plyr")


library(tidyverse)
library(plyr)
library(dplyr)
library(readxl)
#Survival_HeightApril26Blocks7_6Clean <- 
  #read_excel("Oak Project/Data/Survival_HeightApril26Blocks7&6Clean.xlsx", 
                                 #                  sheet = "Input1", skip = 1)
#View(Survival_HeightApril26Blocks7_6Clean)

#names(Survival_HeightApril26Blocks7_6Clean)



oak_data <- list.files(path = "~/GitHub/AdvancedQuantF26/Oak Project/Data",
                       pattern = "*6Clean.xlsx",
                       full.names = TRUE) |> 
lapply(read_excel,
       na = "NA", sheet = "Input1",
       skip = 1) |> 
  bind_rows()

unloadNamespace("plyr")
```

```{r}

oak_data <-
  oak_data |> 
  dplyr::rename(Page_Num = "Page #",
         Date = "MM/DD/YYYY",
         Recorder ="Name",
         Block ="#...4",
         Tag_Num = "#...5",
         Oak_Type ="NR, S, or A",
         Treatment ="AC, OA, NP, DT, or DL",
         OA_Microclimate ="TH, TI, CN",
         Seedling_Height ="(cm)",
         Leaves_Num ="#...10",
         Dead_Leaves_Num ="#...11",
         Leaf_Greenness ="%",
         Germination_Num ="#...13",
         RCD ="(mm)",
         Notes  ="as written on data sheet")
```

```{r}
oct_data <- read_excel("Oak Project/Data/Survival_HeightOct2025Clean.xlsx",
                       sheet = 'Input1',
                                                      skip = 1)
oct_data <- oct_data |> 
  dplyr::rename(Page_Num = "Page #",
                Date = "MM/DD/YYYY",
                Recorder = "Code",
                Block ="#...4",
                Tag_Num = "#...5",
                Oak_Type ="NR, S, or A",
                Treatment ="AC, OA, SA, NP, or DT",
                Seedling_Height ="(cm)...8",
                Leaves_Num ="#...9",
                Base_Seedling_Circumference = "(cm)...10",
                Base_Seedling_Diameter = "(mm)",
                Alive_Dead = "Yes, No, or Missing?",
                Notes = "as written on data sheet")

oak_data_full <- full_join(oak_data, oct_data)

```

```{r}
oak_data_full <- 
  oak_data_full |> 
  mutate(Seedling_Height = as.numeric(Seedling_Height))

#str(Survival_HeightApril26Blocks7_6Clean)

oak_data_full <- 
  oak_data_full |> 
  mutate(Leaves_Num = as.numeric(Leaves_Num))


oak_data_full <- 
  oak_data_full |> 
  mutate(Dead_Leaves_Num = as.numeric(Dead_Leaves_Num))


oak_data_full <- 
  oak_data_full |> 
  mutate(Leaf_Greenness = as.numeric(Leaf_Greenness))

oak_data_full <- 
  oak_data_full |> 
  mutate(Germination_Num = as.numeric(Germination_Num))

oak_data_full <- 
  oak_data_full |> 
  mutate(RCD = as.numeric(RCD))


oak_data_full <- 
  oak_data_full |> 
  mutate(Base_Seedling_Circumference = as.numeric(Base_Seedling_Circumference))

oak_data_full <- 
  oak_data_full |> 
  mutate(Base_Seedling_Diameter = as.numeric(Base_Seedling_Diameter))

oak_data_full <- oak_data_full |> 
  filter(Page_Num != "NA")




oak_data_clean<- oak_data_full |> 
  separate(Tag_Num, into = c("Tag_Num", "Acorn_Num"), sep = "A") |> 
  mutate(Tag_Num = as.numeric(Tag_Num),
         Acorn_Num = as.numeric(Acorn_Num))


oak_data_clean <- oak_data_clean |>
mutate(Dead = case_when(
  Seedling_Height == 0 ~ "Dead",
  Seedling_Height == "NA" ~ "Dead",
  Seedling_Height == NA ~ "Dead",
  Seedling_Height == "NA" & Alive_Dead == "missing" ~ "missing",
Seedling_Height > 0 ~ "Alive"
))

#below dont worry about
         
#Acorn_ID_table <- oak_data_clean |> 
 # select(Tag_Num, Block, Treatment, OA_Microclimate, Date) |> 
  #distinct() |> 
  #arrange(Tag_Num) |> 
  #mutate(Acorn1 = 1, Acorn2 = 2, Acorn3 = 3) |> 
  #pivot_longer(cols = c(Acorn1, Acorn2, Acorn3),
   #            names_to = "Whatever",
    #           values_to = "Acorn_Num") |> 
  #select(-Whatever)

#oak_data_clean |> full_join(Acorn_ID_table)
```

```{r}
#Graph Time!!
oak_data_graph <- oak_data_clean |> 
filter(Seedling_Height != 0) 
  


ggplot(oak_data_graph) +
  aes(x = Seedling_Height) + 
  geom_bar()
#Some of these seedlings appear really high

ggplot(oak_data_graph) +
  aes(x = Leaves_Num) + 
  geom_bar()
#Few data points at 300+ leaves

ggplot(oak_data_graph)+
  aes(x = Leaf_Greenness)+
  geom_bar()

ggplot(oak_data_clean)+
  aes(x = Dead)+
  geom_bar()
#There are a ton of NA's, what does this mean in context to dead or alive?

ggplot(oak_data_clean)+
  aes(x = Seedling_Height, fill = Treatment)+
  geom_bar()

ggplot(oak_data_clean)+
  aes(x = Treatment, y = Seedling_Height, fill = OA_Microclimate)+
  geom_boxplot()

ggplot(oak_data_clean)+
  aes(x = Treatment)+
  geom_bar()



```

- My goal will be to look at Micro climate (DT/AC/OA) and their effect on seedling height and greenness and RCD

- Goals:

  - Focus and isolate DT,AC, and OA treatments (DONE)

  - Acorns (at the end) vs natural recruits (NR) overtime (analyze separately) (DONE)

  - Get rid of all NA's for height and greenness (DONE)

  - ANOVA for your statistical test (Our response is quantitative, or predictor is categorical)

    - For Tuesday, have the data fully cleaned to prepare for the question

- Task 1: Data Cleaning

  - The data is mostly clean, need to edit the data set a bit

```{r}
#Isolating treatments 

Height_Green_RCD <- oak_data_clean |> 
  filter(Treatment == "DT"|
         Treatment == "AC"|
         Treatment == "OA") |> 
  mutate(Date = as.Date(Date, format = "%d.%m.%y"))

str(Height_Green_RCD)
#Removing NAs for height and greenness

Height_Green_RCD <- Height_Green_RCD |> 
  filter(Seedling_Height > 0) |> 
  mutate(Leaf_Greenness = as.numeric(Leaf_Greenness))

#There are some entries that contain a height but not a leaf greenness, and RCD.. etc, will need to split them up into three different dfs
#That also means I will need to seperate them further into Acorns vs New Recruits 
#I will branch between new recruits and acorns first 

HGR_Acorn <- Height_Green_RCD |> 
  filter(Oak_Type == "A")

HGR_Recruit <- Height_Green_RCD |> 
  filter(Oak_Type == "NR")

#Native Recruit Dataset ready to go, acorn dataset may need a bit more

#For acorn: Height has no NAs, while Greenness and RCD have NAs
#Will need to split up the dataset into Greenness and RCD

Acorn_Green <- HGR_Acorn |> 
  filter(!is.na(Leaf_Greenness))

Acorn_RCD <-HGR_Acorn |> 
  filter(!is.na(RCD))


```

- The data is cleaned and ready to go for analysis!!

```{r}
#exploratory analysis (october data removed for first graph)

HGR_Acorn |> 
  filter(Date != "2025-10-21") |> 
ggplot()+
  aes(x = Date, y=Seedling_Height)+
  geom_point()#+
  #geom_line(aes(group =Tag_Num, color = factor(Acorn_Num)))
HGR_Acorn |> 
  #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Seedling_Height)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Planted Acorn Seedling Height")


HGR_Recruit |> 
  #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Seedling_Height)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Natural Recruit Seedling Height")

#Both data are very skewed for height 

Acorn_Green |> 
#filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Leaf_Greenness)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Planted Acorn Leaf Greenness")


HGR_Recruit |> 
 # filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Leaf_Greenness)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Natural Recruit Leaf Greenness")

#Even more skewed for greenness

Acorn_RCD |> 
#filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=RCD)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Planted Acorn RCD")


HGR_Recruit |> 
  filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=RCD)+
  geom_bar()+
  #geom_jitter(color = "red")+
  labs(title = "Natural Recruit RCD")
#RCD is a little better?
#Still really difficult data to work with

HGR_Acorn |> 
  #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=Seedling_Height)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Planted Acorn Seedling Height")

HGR_Recruit |> 
 #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=Seedling_Height)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Natural Recruit Seedling Height")

Acorn_RCD |> 
 # filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=RCD)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Planted Acorn RCD")

HGR_Recruit |> 
 # filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=RCD)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Natural Recruit RCD")


Acorn_Green |> 
  #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=Leaf_Greenness)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Planted Acorn Leaf Greenness")


HGR_Recruit |> 
  #filter(Date != "2025-10-21") |>
  ggplot()+
  aes(x=Treatment, y=Leaf_Greenness)+
  geom_boxplot()+
  geom_jitter(color = "red")+
  labs(title = "Natural Recruit Leaf Greenness")

```

-Consider October as period one, and the rest as period 2

-Natural recruits look at over time (connected dot plots)

-For Thursday, One graph per variable considered presentation ready (one big one to present)

```{r}
#Going to create a series over time graph for natural recruits

HGR_Recruit |> 
  filter(Date != "2025-10-21") |> 
  ggplot()+
  aes(x = Date, y=Seedling_Height)+
  geom_point(size = 2.5)+
geom_line(aes(group =Tag_Num, color = factor(Tag_Num)), linewidth=1)+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 65))+
  labs(x= "Date", y="Seedling Height (cm)", title="Naturally Recruited Acorns' Seedling Height Over Time", color = "Acorn Number" )+
  theme_bw()+
  theme(axis.text.x = element_text(size = 12, color = "black"),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        legend.text = element_text(size = 11, color = "black"),
        legend.title = element_text(size = 12, color = "black"))


#Looks good! Going to do the same with Regular Acorns


HGR_Acorn |> 
filter(Date != "2025-10-21") |> 
  filter(Tag_Num < 250) |> 
  ggplot()+
  aes(x = Date, y=Seedling_Height)+
  geom_point(size = 2.5)+
  geom_line(aes(group =Tag_Num, color = factor(Tag_Num)), linewidth=1)+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 50))+
  labs(x= "Date", y="Seedling Height (cm)", title="Planted Acorns' Seedling Height Over Time (1)", color = "Acorn Number" )+
  theme_bw()+
  theme(axis.text.x = element_text(size = 12, color = "black"),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        legend.text = element_text(size = 11, color = "black"),
        legend.title = element_text(size = 12, color = "black"))


HGR_Acorn |> 
  filter(Date != "2025-10-21") |> 
  filter(Tag_Num > 250) |> 
  ggplot()+
  aes(x = Date, y=Seedling_Height)+
  geom_point(size = 2.5)+
  geom_line(aes(group =Tag_Num, color = factor(Tag_Num)), linewidth=1)+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 20))+
  labs(x= "Date", y="Seedling Height (cm)", title="Planted Acorns' Seedling Height Over Time (2)", color = "Acorn Number" )+
  theme_bw()+
  theme(axis.text.x = element_text(size = 12, color = "black"),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.title.x = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 12, color = "black"),
        legend.text = element_text(size = 11, color = "black"),
        legend.title = element_text(size = 12, color = "black"))
```
